---
title: "Wittenberg University - Master of Science in Analytics"
subtitle: "ANLT 510 - Linear Regression"
author: "Day 1: Course Overview"
date: "`r format(Sys.Date(), format = '%d %b %Y')`"
output: 
  slidy_presentation:
    smart: no
    fig_caption: yes
graphics: yes
params:
  day1_date: "Oct 15"
  day2_date: "Oct 16"
  day3_date: "Nov 12"
---

# Overview

```{r child="resources/R/setup.Rmd"}
```

## Agenda

- This presentation is intended to provide you with a thorough introduction into modeling data using linear regression

- There will be a lot of ideas an concepts to uncover, so there will be several tangents in which we'll break off to make sure you don't get left behind

    + Probability functions
    + Probability distributions
    + Hypothesis testing

## Initial Thoughts

- Linear regression is a simple supervised learning approach

    + Useful tool for predicting a <red>quantitative</red> target responses
    + May seem dull to learn compared to some more modern statistical learning approaches 
    + Is a useful and widely used statistical learning method
    + Comes with some strict assumptions 
    + Serves as a good jumping-off point for other approaches (many newer statistical learning approaches are generalizations or extensions of linear regression)
    + The importance of having a good understanding of linear regression before studying more complex learning methods cannot be overstated

## Prerequisites

- In this presentation we'll work with the [Ames Housing data set](http://jse.amstat.org/v19n3/decock.pdf){target="_blank"}

    + This data set is made available by the `AmesHousing` package (also available in the `modeldata` package)

    + Contains 2930 observations and 80 explanatory variables (23 nominal, 23 ordinal, 14 discrete, and 20 continuous) that are useful for assessing home values

    + Max Kuhn and Julia Silge present a good initial overview of this data set [here](https://www.tmwr.org/ames.html#ames){target="_blank"}

    + We'll use these data to predict the sale price of a house in Ames, IA based on its other characteristics

- As previously mentioned, the initial discovery of relationships is usually done with a training set while a test set is used for evaluating whether the discovered relationships hold

    + A training set is a subset of our data used to discover potentially predictive relationships
    + A test set is a subset of our data used to assess the strength and utility of a predictive relationship

- We'll use a conventional 60% / 40% split where we training our model on 60% of the data and then test the model performance on 40% of the data that is withheld.

```{r prereqs-data, purl=TRUE}
## Load and prep data

# ames data
ames <- AmesHousing::make_ames()

# split data
set.seed(123)
split <- rsample::initial_split(ames,
                                prop = 0.6,
                                strata = "Sale_Price")

ames_train <- rsample::training(split)
ames_test  <- rsample::testing(split)
```

- Before getting started, you'll also need to install/load some R packages and data sets

```{r prereqs-pks-install, purl=TRUE}
install.packages(c("dplyr",
                   "ggplot2",
                   "rsample",
                   "recipes",
                   "vip",
                   "caret",
                   "broom",
                   "plotly",
                   "reshape2",
                   "kableExtra",
                   "mda",
                   "AppliedPredictiveModeling",
                   "earth"))
```


```{r prereqs-pks-load, purl=TRUE}
## Install / load required packages
library(dplyr)
library(ggplot2)
library(rsample)
library(recipes)
library(vip)
library(caret)
library(broom)
library(plotly)
library(reshape2)
library(kableExtra)
library(mda)
library(AppliedPredictiveModeling)
library(earth)

theme_set(theme_bw())
```

## Generalized linear models

- It's important to understand where linear regression fits in context with other models

- The form of a linear regression model is as shown in the expression below, where

    + $Y$ is the target (aka output/response) variable that is a function of $\overline{\beta}}$ and $\overline{X}$
    + $\overline{x}$ is a vector of inputs (the overbar denotes that $X$ is a vector quantity)
    + $\overline{\beta}$ is a vector of weights
    + $\epsilon$ denotes the variation in $Y$ that cannot be accounted for explained by the $X_{0}\beta_{0} + X_{1}\beta_{1}+ \cdots + X_{n}\beta_{n}$

$$
Y = f(\overline{\beta},\overline{X}) = X_{0}\beta_{0} + X_{1}\beta_{1}+ \cdots + X_{n}\beta_{n} + \epsilon
$$

- In words, this expression states that:

    + The values of the output variable $Y$ captured in our data set vary
    + Some of this variation can be accounted for by recognizing the value of $Y$ depends on the value of $X$
    + Building a regression model that includes these inputs may help us understand the relationship between $X$ and $Y$
    + The $\beta$ terms indicate the strength of the relationship between $X$ and $Y$
    + Not all of the variation in $Y$ can be explained by this relationship - this remaining variation is captured by the $\epsilon$ term

- Moreover you should recognize that this model expression has three parts (1 and 2 are easy to see - 3 is not)

    1. A systematic term that expresses how the linear combination ($X_{0}\beta_{0} + X_{1}\beta_{1}+ \cdots + X_{n}\beta_{n}$) impacts the expected value of the target $E[Y|X]$
    
    2. A random term ($\epsilon$) that expresses the amount variance in the target $Y$ that is not explained by systematic term
    
    3. A link function that specifies the link between random and systematic components - expresses the relationship between $E[Y|X]$ and $X_{0}\beta_{0} + X_{1}\beta_{1}+ \cdots + X_{n}\beta_{n}$

- These components are present in all members of the **generalized linear models (GLM)** family

    + Linear and logistic regression are members of this family
    + The table below highlights the members of this family and components for each member

|Model                  | Random      | Link              | Systematic  |
|:----------------------|:------------|:------------------|:------------|
|Linear Regression      | Normal      | Identity          | Continuous  |
|ANOVA                  | Normal      | Identity          | Categorical |
|ANCOVA                 | Normal      | Identity          | Mixed       |
|Logistic Regression    | Binomial    | Logit             | Mixed       |
|Loglinear Regression   | Poisson     | Log               | Categorical |
|Poisson Regression     | Poisson     | Log               | Mixed       |
|Multinomial Regression | Multinomial | Generalized Logit | Mixed       |

- That's a lot math terminology that you may not be familiar with - let's break these ideas down further by reviewing some probability concepts




# Review of Probability and Statistics

## Overview

- You may already know that probability and statistics are related.  This lecture will help you understand this relationship

- To do this we'll define and illustrate the following terms:

- Probability
- Experiment
- Outcome
- Sample space
- Event
- Borel Sigma-Algebra
- Random Variable
- Statistic

- Probability: the branch of mathematics dealing with calculating the likelihood of an event's occurrence, which is expressed as a number $\in [0,1]$

- Statistics: the science of collecting and analyzing numerical data for the purpose of making inferences and/or predictions about a population.

## Probability & Statistics



### Experiment

An experiment is a scientific procedure undertaken to make a discovery, test a hypothesis, or demonstrate a known fact. This is accomplished by analyzing the data generated throughout the experiment.

When you hear the word experiment you may be thinking of scientists in lab coats or remembering back to you experiences in a Chemistry class. While these are accurate representations of an experiment, I'd like for you to think more broadly - think of an experiment simply as a process that generates data that will be used to help make some decision. 

As a data analyst, it's likely that your weren't involved in setting up an experiment.  Instead, you were asked to analyze some data that had already been collected

From this definition we see that we are surrounded by experiments

- The time required for a computer to boot up each morning
- The result of a coin flip
- The result fo rolling a die

Understanding the type of data generated by an experiment and the manner in which the data are collected are crucial to selecting a valid approach  best to model the data   

```{block, echo=F}
IN SOME CASES YOU ARE PART OF AN EXPERIMENT

Need to discuss the work to set up an experiment make sure data supports the decision to be made or hypothesis to be tested

As analyst 

- may be asked to design experiment to collect data used to make decision
- May be asked to use already collected data to determine which decision to make (more common situation)

- Often data is collected without fully understanding the experiments that will be used or the the experiments change over time and the data collected does not provide sufficient infor to answer new questions 
```

For the rest of this lecture we'll use a simple experiment in which we roll a 6-sided die one or more times 

### Outcomes

An outcome is any result that could <u>possibly</u> be observed from an experiment.  If our experiment involved a single roll of a 6-sided die, it is trivial to say that the possible outcomes are $\{1\}$, $\{2\}$, $\{3\}$, $\{4\}$, $\{5\}$, or $\{6\}$. Similarly we understand that an impossible outcome would be any number that is not an integer and any integer that is either less than zero or greater that 7.

Note that each outcome is enclosed within curly brackets 

Suppose, an experiment involves rolling a 6-sided die two times, what would some possible outcomes be?  Many times when I ask students this question the response is "Any number between 2 and 12" - but this is incorrect.  This response assumes that I want you to take the result of the two rolls and add them together. While I may eventually ask you to do this, this is not the same as the possible outcomes 
may as  some possible outcomes would be $\{1,5\}$, $\{3,1\}$, or $\{6,6\}$. An outcome that would not be possible would be $\{7,2\}$.

### Sample space

The sample space refers to the collection of all possible outcomes that could result from an experiment In the literature the sample space is often denoted by the Greek capital letter omega $\Omega$ or a capital script "S" $\mathcal{S}$

### Events

An event is a set of outcomes from an experiment (a subset of the sample space) to which a probability can be assigned. A single outcome may be an element of many different events, and different events in an experiment are usually not equally likely, since they may include very different groups of outcomes.

### Random Variable

Informally, a random variable defined as a variable whose value is random (duh).


### Borel Sigma-Algebra

Represents the mapping of the outcomes to the events

### Statistic
